---
title: "Homework Week 1 DSE5002"
author: "Steven Simonsen"
date: "2024-03-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1
## Assign 10 to the variable x. Assign 5 to the variable y. Assign 20 to the variable z.

```{r}
x <- 10
y <- 5
z <- 20
```

# Exercise 2
## Show that x is less than z but greater than y. Note: your output must be a SINGLE boolean, do not output a boolean for each expression.

```{r}
print(y < x & x < z)
```

# Exercise 3
## Show that x and y do not equal z. Note: your output must be a SINGLE boolean, do not output a boolean for each expression.

```{r}
print(x + y == z)
```

# Exercise 4
## Show that the formula x + 2y = z. Note: your output must be a SINGLE boolean.

```{r}
print(x + 2*y == z)
```


# Exercise 5
## I have created a vector (test_vector) of integers for you. Determine if any of x, y, or z are in the vector. Note: your output must be a SINGLE boolean, do not output a boolean for each expression. test_vector <- c(1,5,11:22)

```{r}
test_vector <- c(1,5,11:22)

print(x | y | z %in% test_vector)
```


# Exercise 6
## Show which value is contained in the test vector. To do this you will need to create an element-wise logical vector using operators. x == vector. Once you have done that you will need to use slicing to return all indices that have matches. Note: your output should be two integers.

```{r}
#Use which function and the OR operator embedded within to check indices 
#of each variable in comparison to the vector. Result = 2, 12
which(x==test_vector | y==test_vector | z==test_vector)
```


